---
title: "Brainconductor demo"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up

We first install the relevant packages, \texttt{brcbase} (which contains
the core code needed for this demo) and \texttt{brcdata} (which contains the
example datasets).

```{r, cache = T}
library(devtools)
devtools::install_github("cdgreenidge/brcbase", ref = "kevin", subdir = "brcbase")
devtools::install_github("linnylin92/brcdata", ref = "kevin")
```

## Basics of \texttt{brcbase}

Let us investigate first \texttt{COBRE\_0040071\_funcSeg}, the functional MRI scan
of subject 0040071 in the COBRE repository. We can look up more information on this
dataset using \texttt{?COBRE\_0040071\_funcSeg}

```{r}
library(brcbase)
library(brcdata)
dat <- brcdata::COBRE_0040071_funcSeg
class(dat)
dat
```

We see some distinctive components of \texttt{dat}, an object of the \texttt{BrcFmri}
class. These are \texttt{data2d}, \texttt{id}, and \texttt{parcellation}. 
```{r}
names(dat)
```

The first component is \texttt{data2d}. In our case, this is a $5 \times 82470$ matrix,
where each column represents a different time index and each row represents a different 
voxel. We call this the ``2d'' representation since it ignores spatial
information, as it is unclear a priori which voxels neighbor which voxels.
